Topic 6.2: Open Data

### 2020-05-06 draft_chapter.Rmd
Edits and additions 3

The drive towards transparency and scrutiny of public information is further enhanced by an increasing need for replication - a hallmark for open science research practice. Opening up data sets used in criminology publications and wider social sciences fosters and facilitates a culture of replication [@pridemore2018replication]. As such, open data fits into the wider discourse around transparency, open review and open scrutiny, which (in time) may even become a policy requirement irrespective of the research field or the source of funding [@vuong2017open]. Increasingly, researchers are making use of online repositories for their papers (e.g. www.osf.io), data (e.g. www.ukdataservice.ac.uk/deposit-data) and code (e.g. www.github.com), all in the name of transparency and replication. These are guidelines and tools worth considering, not just when conducting analysis on secondary data sources, but also collecting and sharing your own data.

@@

Opening up data generated by the private sector represents a significant challenge, largely due to the proprierary value to its creators [@kitchin2014data]. The aims and objectives of private companies, who have a duty to shareholders and operate in a competitive market environment, differ considerably from local and central government. That said, some datasets are released openly by private sector organisations, albiet often only a subset of what would be available as a paying customer. The website for ArcGIS, a piece of software maintained by the Environmental Systems Research Institute (ESRI), a private company, host a number of open geographic datasets on their online cloud platform (www.hub.arcgis.com/search). Indeed, many papers have made use of data collected or distributed by private organisations to explore crime and place, such as Google Street View [@langton2017residential] and Twitter [@malleson2015impact]. In this book you will learn to use the free Twitter API for such purposes in criminology (see Section 3, Topic 7).

@@

Finally, there is a specific group of open data sources that collate data collectively, generated by large groups of individuals who do not specifically belong to any organisation, but instead work together on a collaborative project. These are crowdsourced data sets. Examples include Wikipedia (www.wikipedia.org), an online encyclopedia where anyone can contribute, and Flickr (www.flickr.com) an online photo gallery where people can upload and tag their photos. Often, the organisations who maintain and monitor these data collection activities are charities or non-govermental organisations that do not operate for profit, but instead provide some form of social good. An example is the online reporting platform FixMyStreet (www.fixmystreet.com), where people can report problems such as instances of graffiti, vandalism or environment issues. The data has been used to explore signal crimes theory, for instance, offering insight into people's experiences with incivilities [@solymosi2018crowdsourcing]. Another chapter in this book (see Section 3, Topic 6) discusses the merits and pitfalls of crowdsourced data, and what to watch out for when analysing open data of this type.

@@

- Accessing open data using Application Programming Interfaces (APIs) and direct downloads.
- Cleaning and wrangling data for analysis.
- Visualising open spatial data.
- Comparing different sources of open data.
- Engaging with the critical questions outline above to better understand open data.

@@

In demonstrating these skills, we will be answering two key research questions. The first is a substantive empirical investigation into crime in and around London Underground stations, in the spirit of existing research into transport crime (see newton2014above). Studies in Sweden, for instance, have shown the importance of environmental and neighbourhood characteristics in determining crime concentrations at underground stations in Stockholm, along with the positioning of stations on the line [@ceccato2013security]. Investigations have also taken place into crime concentrations at bus stops in Los Angeles [@loukaitou1999hot] and the impact of intensive policing along bus corridors in Merseyside, England [@newton2004crime]. As auch, we can pose the following research question:

- RQ1: To what extent does crime cluster in and around London underground stations?

@@

First, we will go through how to access the data used in the exercises: an important skill when seeking to use open data.

@@

We will be using two different types of open data in this exercise, public sector data (crime data and transport authority data) and crowdsourced data (from Open Street Map). To access them we will use three different methods: 1) Direct download, 2) Direct request to an API, 3) Request to an API using a wrapper. This will give you few different ideas about how you might go about accessing other open data sets relevant to your research.

@@

The simplest way that open data is often made available is through direct download from a website. This is the case for open police data made available in the United Kingdom through the [data.police.uk](https://data.police.uk/) webportal. To access this data, simply type the url into any web browser ( [https://data.police.uk/](https://data.police.uk/) ), and follow the guidance to acquire data. In the case of police.uk you [select the tab which says "Data"](https://data.police.uk/data/), choose your "Date range:" (ie for what months you would like to download crime data for), choose the police forces you want crime data from, choose some options (whether you want to include crime data, outcomes data, and/or stop and search data) and click on 'Generate File' to download.

Once you download this data you will have to save this locally on your computer. We have saved ours in a subfolder called "data" and will be loading the data into R directly from there.

Some data downloads may come bundled with a data dictionary, explaining important information about your data set, which you will need to answer the critical questions outlined in previous sections. In other cases this might be found online. This is the case with the police.uk dataset, where the data dictionary can be found on the ["About" page](https://data.police.uk/about/). Make sure you always read through this correctly so you can answer your critical questions and fully understand your data set.

@@

An application programming interface (API) is a tool which defines an interface for a programme to interact with a software component or system, for example, what sort of requests or calls can be made to it, and how these can be made. We are specifically using the term APIs to denote tools created by an open data provider to provide access to different subsets of their content. APIs facilitate scripted and programmatic extraction of content, as permitted by the API provider [@olmedilla2016harvesting]. APIs can take many different forms and be of varying quality and usefulness [@foster2016big]. For the purposes of accessing open data from the web, we are specifically talking about **RESTful APIs**. REST stands for Representational State Transfer. These APIs work directly over the web, meaning that we as users can play with the API with relative ease to understand how it works [@foster2016big].

Here we will use the example of the British transit authority Transport for London (TfL) to practice constructing HTTP requests via their API, and then parsing the data that are returned. Much like how downloading data directly requires access to a data dictionary to make sense of the data and answer our critical questions, we must find a similar document to understand how the API we plan on using works. In the case of the TfL API this is provided via the [TfL Open Data page for their unified API](https://tfl.gov.uk/info-for/open-data-users/unified-api) and the [documentation page](https://api-portal.tfl.gov.uk/docs). We will engage more with this in the exercises.

@@

Often developers who work with APIs will share their code for doing so, and release them in the form of a package or module. This is called a **wrapper** because it's code that wraps the API in a nice package, making it easier for future users to access. The wrapper can take many forms, it could be a Python module, or an R package, or it can even be a web interface that provides a graphical user interface (GUI) for accessing the API in question.

To demonstrate this, we will be accessing data from Open Street Map (OSM). OpenStreetMap is a crowdsourced open data set of spatial information built by a community of mappers that contribute and maintain data about all sorts of environmental features, such as roads, trails, cafes, railway stations, and much more, all over the world. The result is a map rich in local knowledge, which prvides valuable information about places and their features.

OSM has two types of wrappers available for its API, a web-based GUI called [Overpass Turbo](https://overpass-turbo.eu/), and an R package called `osmdata`. We will explore using both.

<!-- ### Open Street Map -->

<!-- #### What is Open Street Map? -->

<!-- - Brief history. -->
<!-- - Motivations for and advantages of an open web mapping platform. -->
<!-- - Primary features (keys, values, elements). -->
<!-- - Contributing to the data yourself. -->

<!-- ### data.police.uk -->

<!-- ### Transport for London Open Data Portal -->

<!-- ### Downloading Open Street Map Data -->

<!-- - Define and explain APIs. -->
<!-- - For OSM, describe overpass queries and overpass-turbo. -->
<!-- - Query example (brief). -->

<!-- ### Using Open Street Map data in R -->

<!-- - Made easy through the `osmdata` package in R. -->
<!-- - More straightforward than overpass-turbo. -->
<!-- - Compatible with both `sp` and `sf` classes of spatial data. -->
<!-- - We focus on `sf` as its compatible with the `tidyverse`. -->
<!-- - Outline the key documentation and references. -->

@@

Let's get started! Getting back to our problem at hand, crime at bus stops, we want to know how crimes and bus stops interact...
<!-- clear up what research q we can actually answer....!! -->

@@

```{r}
library(jsonlite)
library(osmdata)
library(ggplot2)
library(sf)
library(readr)
library(dplyr)
library(tidyr)
```

Now let's focus on acquiring our different data sets

### Direct Dowload of crime data

<!-- IDK how much detail we actually need here, probably not loads, about downloading extracting and importing police.uk data -->

@@

Our first step is to read through the . The tells us what sort of data is possible to acquire. We know that currently we want to find the location of bus stops, and any information that is available from TfL about these bus stops. In the documentation is where we find specifically *how* we can request this.

Specifically, we can scroll down to where we find "". Next to this we see the example URL: `https://api.tfl.gov.uk/line/24/route/sequence/outbound` This example demonstrates how we can request this information, and gives us a template to alter, but also allows us to take a peak into the result this call will return.

To see, try to take the above URL, and copy and paste it into your browser window. What do you see? It should be something like this:

```{r tfl, out.width = "\\textwidth", fig.cap="Data from TfL"}
knitr::include_graphics("img/tfl_screenshot.png")
```

This contains the information we requested (bus route 24 stops outbound) in a format called **JSON**, which stands for JavaScript Object Notation. JSON is an open standard way of storing and exchanging data, and will most likely be the format in which data are returned from most API calls.

Great, but how do we get this into R? Well this is where the `rjson` package is handy, and specifically we can use the `fromJSON()` function to parse this JSON p=object, which we can read directly from the URL into R using the `readLines()` function (base R). So, if we wanted the results of the bus route 24, we would type:

```{r getrt24, eval = F}
api_call <- fromJSON(readLines("https://api.tfl.gov.uk/line/24/route/sequence/outbound"))
```

This gives us an object (`api_call`) which contains all the information returned by the TfL API. Now JSON is slightly different to dataframes which we are more used to dealing with, in that there is nested information

```{r conferttodf, eval = F}
json_data_frame <- as.data.frame(api_call)
```

>>>>

The drive towards transparency and scrutiny of public information is further enhanced by an increasing need for replication - a hallmark for open science research practice. Opening up data sets used in criminology publications and wider social sciences fosters and facilitates a culture of replication [@pridemore2018replication]. As such, open data fits into the wider discourse around transparency, open review and open scrutiny, which (in time) may even become a policy requirement irrespective of the research field or the source of funding [@vuong2017open]. Increasingly, researchers are making use of online repositories for their papers (e.g. https://osf.io/), data (e.g. https://www.ukdataservice.ac.uk/deposit-data) and code (e.g. https://github.com/), all in the name of transparency and replication. These are guidelines and tools worth considering, not just when conducting analysis on secondary data sources, but also collecting and sharing your own data.

@@

Opening up data generated by the private sector represents a significant challenge, largely due to the proprierary value to its creators [@kitchin2014data]. The aims and objectives of private companies, who have a duty to shareholders and operate in a competitive market environment, differ considerably from local and central government. That said, some datasets are released openly by private sector organisations, albiet often only a subset of what would be available as a paying customer. The website for ArcGIS, a piece of software maintained by the Environmental Systems Research Institute (ESRI), a private company, host a number of open geographic datasets on their online cloud platform (https://hub.arcgis.com/search). Indeed, many papers have made use of data collected or distributed by private organisations to explore crime and place, such as Google Street View [@langton2017residential] and Twitter [@malleson2015impact]. In this book you will learn to use the free Twitter API for such purposes in criminology (see Section 3, Topic 7).

@@

Finally, there is a specific group of open data sources that collate data collectively, generated by large groups of individuals who do not specifically belong to any organisation, but instead work together on a collaborative project. These are crowdsourced data sets. Examples include Wikipedia (https://www.wikipedia.org/), an online encyclopedia where anyone can contribute, and Flickr (https://www.flickr.com/) an online photo gallery where people can upload and tag their photos. Often, the organisations who maintain and monitor these data collection activities are charities or non-govermental organisations that do not operate for profit, but instead provide some form of social good. An example is the online reporting platform FixMyStreet (https://www.fixmystreet.com/), where people can report problems such as instances of graffiti, vandalism or environment issues. The data has been used to explore signal crimes theory, for instance, offering insight into people's experiences with incivilities [@solymosi2018crowdsourcing]. Another chapter in this book (see Section 3, Topic 6) discusses the merits and pitfalls of crowdsourced data, and what to watch out for when analysing open data of this type.

@@

- Accessing open data using direct downloads and Application Programming Interfaces (APIs).
- Cleaning and wrangling data for analysis.
- Visualising open spatial data.
- Comparing different sources of open data.
- Engaging with the critical questions outline above to better understand open data.

@@

In demonstrating these skills, we will be answering two key research questions. The first is a substantive empirical investigation into crime in and around London Underground stations, in the spirit of existing research into transport crime (see newton2014above). Studies in Sweden, for instance, have shown the importance of environmental and neighbourhood characteristics in determining crime concentrations at underground stations in Stockholm, along with the positioning of stations on the line [@ceccato2013security]. Investigations have also taken place into crime concentrations at bus stops in Los Angeles [@loukaitou1999hot] and the impact of intensive policing along bus corridors in Merseyside, England [@newton2004crime], amongst others. As auch, we can pose the following research question:

- RQ1: To what extent does crime cluster in and around London Underground stations?

@@

First, we will go through how to access the open data used in the exercises.

@@

We will be using two different types of open data in this exercise: public sector data (police recorded crime data and local transport authority data) and crowdsourced data (from Open Street Map). To access them we will use three different methods, namely, (1) direct download, (2) direct request to an API, 3) request to an API using a wrapper. This will give you few different ideas about how you might go about accessing other open data sets relevant to your research. Locating, identifying and learning to access open data is a skill in itself.

@@

The simplest way that open data can be made available is through direct download from a website. This is the case for open police recorded crime data in England and Wales. It can be accessed and downloaded using an online web portal (https://data.police.uk/) under Open Government Licence. Using the "Data" tab, anyone from researchers to members of the public can select the time period and police force of interest, and the type of information required (e.g. crimes, stop and search, outcomes), and then "Generate file" to create a download link. Data is downloaded in  comma-separated values format (.csv file extension). The website also contains detailed information about data quality, downloads and tools under the "About" tab, which is particularly useful when considering the critical questions posed earlier (where, why, who, what, when). For this exercise, we are going to use British Transport Police data, a force which operates on railways and light-rail systems across the country, for the month of January in 2020. Select the time period using the dropdown menus, the force using the tickboxes, and then generate and download the file. We save this file locally to a subfolder named "data", and during the exercise, the data is loaded into R from this folder.

@@

Another way of downloading open data is through an application programming interface (API). This is a tool which defines an interface for a programme to interact with a software component or system, for example, what sort of requests or calls can be made to it, and how these can be made. Here, we are using the term 'API' to denote tools created by an open data provider to provide access to different subsets of their content. APIs facilitate scripted and programmatic extraction of content, as permitted by the API provider [@olmedilla2016harvesting]. APIs can take many different forms and be of varying quality and usefulness [@foster2016big]. For the purposes of accessing open data from the web, we are specifically talking about _RESTful_ APIs. The 'REST' stands for Representational State Transfer. These APIs work directly over the web, which means users can play with the API with relative ease in order to understand how it works [@foster2016big].

Here, we will use the example of a local government organisation responsible for transport, Transport for London (TfL). One of the main responsibilities of TfL is to oversee various rail networks in Greater London, including the London Underground. We will use open data from TfL to practice how to construct HTTP (i.e. web link) requests via their API, and then parse the data which is returned. Much like how downloading data directly requires access to a data dictionary to make sense of the data (e.g. understanding its origins and flaws), we must find a similar document to understand how the API we plan on using works. Details about the TfL API are provided through their open data page (https://tfl.gov.uk/info-for/open-data-users/unified-api), which explains what data is available, and how the API is designed. Usefully, TfL provide a documentation page with examples of how you can use the HTTP links to access make calls to the API (https://api-portal.tfl.gov.uk/docs). We will actually make these calls from within the R environment using a 'wrapper', detailed below, but it might be useful (and interesting) to examine how basic queries are constructed using this documentation.

@@

Often, developers who work with APIs will share their code, and release them in the form of a package or module. This is called a _wrapper_ because it uses code that 'wraps' around the API to make it a neater, more usable package. Wrappers remove (or at least lower) many of the obstacles to accessing open data noted earlier, particularly for people from a non-technical background. The wrapper can take many forms, such as a Python module, or an R package. It could even be a web interface that provides a graphical user interface (GUI) for accessing the API in question.

To demonstrate this, we will be accessing data from Open Street Map, a database of geospatial information built by a community of mappers, enthusiasts and members of the public, who contribute and maintain data about all sorts of environmental features, such as roads, green spaces, restaurants and railway stations, amongst many other things, all over the world. As such, it is a prime example of 'crowdsourced' open data. You an view the information contributed to Open Street Map using their online mapping platform (https://www.openstreetmap.org/). The result of people's contributions is database of spatial information rich in local knowledge which provides invaluable information about places and their features, without being subject to strict terms on usage. Open Street Map has two types of wrappers available for its API, a web-based GUI called Overpass Turbo (https://overpass-turbo.eu/), and an R package called `osmdata`. We will explore the APIs using both, but focus on their usage from within R.

@@

So, let's get started. Recall that are interested in examining the extent to which crime concentrates in and around London Underground stations, but also the contrasts between the open data available via TfL and Open Street Map, as public sector and crowdsourced open data, respectively.

@@

```{r}
library(jsonlite) 
library(osmdata)
library(ggplot2)
library(sf)
library(readr)
library(dplyr)
library(tidyr)
library(purrr)
library(patchwork)
```

@@

Reading through the TfL API documentation mentioned earlier (https://api-portal.tfl.gov.uk/docs), you will get a feel for the kind of data available. To draw conclusions about the degree to which crime occurs in and around London Underground stations, we first need spatial information so we can plot station locations on a map. In the documentation there in fact an 'API area', which offer some guidance on the types of data available, called _Stops_. We can see that this include information on bus stops by route (which are numbered e.g. 24), and underground stations by line (which are named e.g. "Bakerloo"). The example for the underground stations is a URL: https://api.tfl.gov.uk/line/bakerloo/stoppoints. This has quite an intuitive structure, since we first want to access the TfL API, look for data by line, select the name of the line we are interested in, and finally pull unformation about where stations stops. In fact, you can simply copy and paste this URL into your internet browser, and it should look something like Figure 1.

```{r tfl, out.width = "\\textwidth", fig.cap="Figure 1: example underground stations data from Transport for London", echo = F}
knitr::include_graphics("img/tfl_screenshot.png")
```

Although this contains the information we requested, it is in a format called _JSON_. This stands for JavaScript Object Notation. JSON is an open standard way of storing and exchanging data, and will most likely be the format in which data are returned from most API calls. As you can see, it is intuitive to read in its current format. In R, we can use the `fromJSON()` function from the `rjson` package to parse all this information into a data frame (with rows and columns). It will then become more familiar and usable. All we need to do is input the URL from the TfL API into the `txt` argument of the function, with a bit of help from `readLines()` in base R which is used to read text from URLs.

To keep things focused, let's request data about stop points on the Northern Line. Note how this is simply an amended version of the example provided by TfL in the API documentation.

```{r getapi, eval = F}
api_call <- fromJSON(readLines("https://api.tfl.gov.uk/line/northern/stoppoints"))
```

This gives us an object (`api_call`) which contains all the information returned by the TfL API. JSON is slightly different to tradtional data frames with rows and columns, which we are probably quite familiar with, because the data is nested. For instance, `api_call` is classed as a data frame, but upon viewing the object using `View(api_call)` you will notice that there are lists nested within it. Another way of checking this is looping the `class()` function through all the columns in the data frame using `lapply(api_call, class)`. This demonstrates an important challenge faced by researchers when using open data, because dealing with data in this format can be messy and complicated.

Fortunately for us, R is more than capable of dealing with nested data. Moreover, we can see that the information we are interested in (namely, the location of underground stations) has already been successfully parsed into columns called **lon** (longitude) and **lat** (latitude), along with an identication column of the stop names called **commonNanme**. We can extract these columns, and create a spatial `sf` object using the coordinates, all in one chunk of code, using the piping operator `%>%` in the `dplyr` package. Note that we define the Coordinate Reference System (CRS) which the data has come in, and then instantly transform it into something more appropriate (a projected CRS known as the British National Grid).

```{r apispatial, eval = F}
tfl_north_sf <- api_call %>% 
  select(commonName, lat, lon) %>% 
  st_as_sf(coords = c(x = "lon", y = "lat"), crs = 4326) %>% 
  st_transform(27700)
```
